Artificial ecosystem selection reveals relationships between microbiome composition and ecosystem function

### 2022-11-21 manuscript/manuscript.Rmd
rename and reorganize stuff

^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA

\* Correspondence: Andrew H. Morris, amorris3@uoregon.edu

Competing Interests: We declare we have no competing interests.

```{r options, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

@@

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
# library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
# library(ggthemes)
library(captioner)
library(knitr)
library(ggtext)
figures <- captioner()
tables <- captioner(prefix = "Table")
```

```{r load-data}
heritability <- read_tsv('../Output/heritability.tsv')
fluxes <- read_tsv('../Output/fluxes.tsv')

load(file = '../Output/decontam.RData')

# CH4 models
load(file = '../Output/response_model.RData')
load(file = '../Output/response_plot.RData')
load(file = '../Output/herit_model.RData')
load(file = '../Output/herit_plot.RData')

# 16S models
load(file = '../Output/richness_models.RData')
load(file = '../Output/phyloseq.RData')
load(file = '../Output/beta_model.RData')
load(file = '../Output/beta_fig.RData')
load(file = '../Output/da_corncob.RData')
load(file = '../Output/da_fig.RData')
load(file = '../Output/dbrda.RData')

source('../R/functions.R')

@@

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know a priori which
members of a microbiome influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per round over five rounds of selection. In addition, we
estimated that `r myround((coefficients(herit_model)[2] +
coefficients(herit_model)[4])*100, 1)`\% of the variation in methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; rather microbes that
responded to selection included members of `r length(da_fams)` families not
known to oxidize methane, such as
*Fimbriimonadaceae*, *Cytophagaceae*, and *Diplorickettsiaceae*. This result is
in contrast to the typical assumption that the rate of an ecosystem function is
limited by the final step in the associated microbial pathway. Our study
demonstrates that variation in microbiome composition can contribute to
variation in the rate of ecosystem function and that manipulating microbiome composition
could be a viable strategy for managing ecosystem functions.

@@

Microbiomes mediate a variety of important functions in ecosystems, and there
is great interest in understanding how attributes of microbiomes may influence
variation in ecosystem functions [@conrad1996; @schimel1998; @crowther2019].
Biodiversity-ecosystem function relationships have been documented for a
variety of macroorganismal communities. For example, plant species richness
correlates with productivity and marine community diversity correlates with
functions such as production, consumption, and nutrient release
[@cardinale2006; @hooper2012; @gamfeldt2015]. While there is some evidence that
variation in microbiomes can contribute to variation in ecosystem functions,
describing a more detailed mapping between the microbiome and ecosystem
function has been elusive [@conrad1996; @rocca2015; @graham2016].

There are several reasons for this difficulty. One reason is that it is
challenging to manipulate microbiome composition and diversity directly.
Experimental manipulations of community composition and diversity have been
central to documenting the relationships between ecosystem functions and the
biodiversity of plants and animals [@hooper2012], but this is seldom possible
with microbiomes. Most microorganisms cannot yet be cultured and most
available methods for manipulation of microbiomes (e.g., antibiotics, dilution,
etc.) are not sufficiently specific to manipulate individual microbial
lineages. Microbial ecologists are thus often restricted to using comparative
methods, in which microbiome composition and diversity is surveyed via
environmental DNA and then correlated with the rate of an ecosystem function. 
However, microbiome attributes add an average of only 8% of variance explained
to ecosystem models [@graham2016].

In an
attempt to increase the likelihood of detecting such a correlation, microbial
ecologists may focus their survey efforts on a particular group of microbes
thought to be involved in the ecosystem function of interest. For example, many
studies have looked for relationships between an ecosystem function and the
abundance or diversity of a microbial gene that codes for an enzyme or
enzyme-subunit known to be involved in a particular function. But this approach
assumes that one can identify the “right” marker gene a priori which, given how
little is understood about microbial diversity, seems highly unlikely. And for
most functions in most ecosystems there is no correlation between the abundance
of a marker gene and the rate of the corresponding process [@rocca2015].

Another likely reason that microbiome-ecosystem function relationships have
been difficult to document is that microbiomes often covary with other drivers
of variation in ecosystem functions, such as environmental conditions or
resources. Such covariation can make it impossible to determine whether
microbiomes directly contribute to functional variation. Understanding the
degree to which microbiomes directly influence functional variation is crucial
if scientists are to develop microbiome-focused methods for altering or
maintaining ecosystem functions.

The problem of connecting microbiome attributes to the rate of an ecosystem
function is analogous to the problem of connecting genomic variation to
phenotypic traits in organisms, and solutions to this problem may be found by
exploring this analogy [@morris2020]. For example, one approach to the problem
of genotype-phenotype mapping is to perform artificial selection on a trait of
interest and identify genes that respond to selection. Such genes are likely to
be causally linked to the trait under selection. An analogous approach for
exploring microbiome-ecosystem function relationships is gaining acceptance
among microbiologists [@swenson2000; @blouin2015; @panke-buisse2015; @sanchez2021].
Microbiomes have been shown to respond to selection at the ecosystem-trait
level [@swenson2000], and some ecosystem-scale microbiome traits have been
shown to be transferable through microbiome inoculations [@panke-buisse2015].
Applying artificial selection to whole microbiomes could be a novel approach
for quantifying how much of the variation in the rate of an ecosystem function
could be attributed independently to microbiome variation [@goodnight2000; @blouin2015]. 
By "microbiome variation," we mean aspects of microbial diversity that are
routinely measured by microbial ecologists, such as the relative
abundance of taxonomic groups inferred from amplicon sequence variants. 
In
addition, this approach could be used to identify microbiome attributes, such
as genes or taxa, associated with the rate of an ecosystem function, although,
to our knowledge this has not previously been accomplished [@arias-sanchez2019].
Such an approach has the potential to expand our ability to more
accurately model microbial ecosystem-function relationships and allow us to
manage ecosystems for particular outcomes [@fuhrman2009; @schimel1998].

We use this approach to estimate how much of the variation in an ecosystem
function - the flux of CH~4~ gas between the soil and atmosphere - can be
attributed to variation in the microbiome and to identify microbial markers
associated with this ecosystem function.  Soil CH~4~ oxidation is a suitable
function for this study because there is evidence that it may vary with
microbiome attributes. For example, variation in CH~4~ emissions in arctic
permafrost is correlated with the transcriptional activity of certain
methanogens [@freitag2009; @freitag2010]. In addition, soil microbiome
phylogenetic variation is a strong predictor of CH~4~ oxidation in forests and
pastures of the Brazilian Amazon and the Congo Basin [@meyer2020; @meyer2020a]. Finally, methanogenesis and
methanotrophy are two of the most deeply conserved microbial physiologies and
are represented in a narrow range of taxa and so the taxonomic composition of
the microbiome is more likely to be associated with the rate of CH~4~ flux than
other broader or more shallowly conserved functions [@schimel1998;
@martiny2013]. This suggests that microbiome variation might drive variation in
CH~4~ oxidation in soils.

In this study, we wanted to evaluate whether variation in the microbiome
contributes to variation in the rate of ecosystem functions and to identify
microbiome attributes that might explain the mapping between microbiome and
ecosystem function. To begin to address this problem, we used artificial
ecosystem selection on soil microbiomes as a model system.  Using this
experimental model, we answered the following questions: Does variation in the
relative abundance of microbial taxa contribute to variation in soil methane
oxidation rate in the lab? How much of the variation in soil methane oxidation
rate can be attributed to the microbiome? What attributes of the microbiome
regulate variation in methane oxidation rate? And do these attributes match our
assumptions about what regulates methane oxidation rate?  We applied artificial
ecosystem selection on soil microbiomes by selecting on the ecosystem-scale
CH~4~ oxidation rate. We estimated how much of the variation in CH~4~ oxidation
rate was associated with variation in the microbiome by calculating the slope
of the regression between donor and recipient communities. To identify which
microorganisms could be markers of CH~4~ oxidation rate, we compared the
composition of the artificially selected microbiomes to a control set of
microbiomes with random selection. We then evaluated whether these markers met
our underlying assumptions about which taxa limit the rate of ecosystem
function.

@@

We performed an artificial ecosystem selection experiment [*sensu* @swenson2000]
by passaging replicate soil microbiomes. The trait we selected on was
CH~4~ oxidation rate. This experiment had two selection lines with twelve jars
each for a total of twenty-four jars per passage. One line was a positive
selection line where the two or three jars with the highest CH~4~ oxidation rate
were chosen to inoculate the next set of positive jars. The other line was a
neutral selection line where an equal number of jars as the positive line were
chosen at random to inoculate the next set of neutral jars. The number of jars
chosen was based on the distribution of fluxes among the positive jars, i.e., we
chose the top three jars unless only two jars had considerably greater CH~4~ oxidation
rates based on visual inspection of histograms. The experiment was carried out
over five passages until a significant divergence in functional rates was 
observed between the two selection treatments.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting mix was
Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). The jars were
then homogenized, capped, and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the
two treatment groups, twenty-four jars were created in an identical manner and
then randomly assigned to either the positive or neutral selection line. Jars
were flushed and respiked twice per week to maintain aerobic conditions and
elevated CH~4~ concentrations and were incubated at ambient temperature for
approximately three weeks. Methane oxidation rates were determined at the end
of the incubation period. For the positive treatment, the three jars with the
greatest CH~4~ oxidation rates were chosen to inoculate the next generation.
For the neutral treatment, three jars were randomly selected to inoculate the
next generation. For each treatment, the selected jars were homogenized and 5 g
of the homogenized soil was used to inoculate the next set of jars, which
represents a 10% subsampling. The next set of jars were created in an identical
manner to the first generation with fresh autoclaved potting mix and the same
moisture and CH~4~ content.

@@

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from
each jar immediately after spiking and then at time points 3, 6, 24, and 48
hours for a 5-point curve. Samples were immediately injected into a SRI model
8610C gas chromatograph equipped with a flame ionization detector (SRI
Instruments, Torrance, CA, USA) to determine the headspace CH~4~ concentration.
Fluxes were calculated from a first-order exponential decay function as decay
constant $k$ with units day^-1^. Oxidation rates are presented as the additive
inverse of $k$ (i.e., $-k$) so that a more positive value represents a greater
oxidation rate. The jars selected for the positive treatment in passage 2 had
the lowest CH~4~ oxidation rate of the twelve jars due to a calculation error
in the CH~4~ oxidation rate. Despite this error, we still observed a strong
response to selection demonstrating the power of this technique. In addition,
this likely reintroduced diversity lost during the selection process,
potentially increasing the variation available to be selected upon, as
is recommended for microbiome selection experiments [@sanchez2021].

@@

Statistical analyses were performed in 'R' [@rcoreteam2018]. To determine
whether there was a significant change in CH~4~ oxidation rate as a
response to selection, we tested a difference in slopes between the positive
and neutral selection lines. Residuals did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed prior to analysis. First, we tested the difference of slopes between
the positive and neutral treatments by
fitting two nested models with and without the interaction between treatment
and passage using 'lm'. We compared these
models using an F-test with the 'anova' function. We
then fit the full model with the interaction term to estimate the slopes for each treatment, which
represented the change in CH~4~ oxidation rate per passage as a response to
selection.

We estimated the proportion of variance in CH~4~ oxidation rate due to variation in
the microbiome as the
regression of recipient on donor [@falconer1996]. The mid-donor was the
mean for all three selected jars and the mid-recipient was the mean methane oxidation 
rate for the twelve jars produced by those three donors. First, we tested if there was an effect of treatment by comparing nested models with and without treatment using an F-test with the 'anova' function. We then fit the full model to estimate the
slope of the relationship between recipient and donor for the Positive and Neutral treatments.

@@

To identify taxa that responded to selection on CH~4~ oxidation rate, we tested
differential abundance between the two treatments in passage 5. We first
grouped ASVs at the family level keeping ASVs that lacked a family-level taxonomic
assignment by grouping them at a higher level. We
then subset the samples in Passage 5 and removed all families with a prevalence
of less than 10\% in either treatment. We used three methods for testing differential abundance: ANCOM-II, ALDEx2, and CORNCOB. We then
identified the consensus taxa that were significant
with all three tests and plotted their relative abundances. For ANCOM-II, we used the 'ancom' function in the 'ANCOM-BC' package with a cutoff of W = 0.7 [@kaul2017; @lin2020]. For ALDEx2, we used the 'aldex' function in the 'ALDEx2' package with Welch's t-test and we used an effect size of
1 as our significance threshold [@fernandes2014]. Finally, we used CORNCOB with the 'differentialTest' function in the 'corncob' package with the Wald test and without bootstrapping [@martin2020]. Lastly, to test differentially abundant methanotrophs, we subset all ASVs within methanotrophic families and tested their differential abundance aggregated at the family and genus level using 'corncob'.

@@

We observed a response to selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`; difference of slopes:
F~`r compare_response_mods[2,4]`,`r compare_response_mods[2,2]`~ = `r myround(compare_response_mods[2,5], 2)`, p = `r myround(compare_response_mods[2,6], 2)`). At the start of the experiment, the Positive
treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\% lower than the
Neutral treatment `r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate of the Neutral
treatment over the five passages `r print_lm(response_model, 2)`. By contrast,
the Positive treatment had a `r myround(per_slope, 1)`\% increase in CH~4~
oxidation rate per passage `r print_lm(response_model, 4)`.

We estimated the proportion of variation in methane oxidation due to variation
in the microbiome as the regression of mid-recipient on mid-donor 
(`r figures("herit", display = "cite")`).  Recipient CH~4~ oxidation rates were
correlated with donor CH~4~ oxidation rates in both the Positive treatment
`r paste0('(', 'slope', ' = ', myround(tidy(herit_model)[2, 2] + tidy(herit_model)[4, 2], 2), ', SE = ', myround(tidy(herit_model)[4, 3], 2), ', t = ', myround(tidy(herit_model)[4, 4], 2), ', p = ', myround(tidy(herit_model)[4, 5], 2), ')')` and 
the Neutral treatment `r print_lm(herit_model, 2)`.

@@

While it is well understood that the microbiome mediates a variety of ecosystem
functions, a fundamental question in microbial ecology is whether we need to
understand variation in the composition of microbiomes to understand variation
in the rate of ecosystem functions [@conrad1996; @schimel1998]. To address this
question, we performed artificial ecosystem selection on soil microbiomes by
selecting for ecosystems with high methane oxidation rates. We found that there
was a significant increase in methane oxidation rate in the Positive selection
treatment relative to the Neutral selection treatment. This suggests that 
variation in the microbiome can generate variation in ecosystem
functions independent of the environment.

We wanted to estimate how much of the variation in ecosystem function can be
attributed to variation in the microbiome. This problem is analogous to the
problem in quantitative genetics of estimating how much of the variation in an
organismal trait is due variation in genotype, i.e., heritability. Organismal
biologists often estimate the percent heritability of a trait as the slope of the
regression of offspring phenotype
on parental phenotype [@falconer1996]. A correlation between the trait values 
of parents and offspring indicates that the trait is at least partially 
genetically determined. Using artificial
ecosystem selection we can estimate the variation in an ecosystem trait due to
variation in the microbiome by comparing the trait values of the "offspring" or
recipient jars in one generation to the trait value of the "parental" or donor
jars chosen for selection in the previous generation. This "ecosystem
heritability" then quantifies what percentage of the variation in ecosystem function
is due to variation in the microbiome [@goodnight2000; @blouin2015].

>>>>

^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA  
\* Correspondence: Andrew H. Morris, amorris3\@uoregon.edu  
Competing Interests: We declare we have no competing interests.

```{r options, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_knit$set(root.dir=normalizePath('../'))

@@

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
# library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
# library(ggthemes)
library(captioner)
library(knitr)
library(ggtext)
figures <- captioner()
tables <- captioner(prefix = "Table")

source('R/functions.R')
```

```{r load-data}
heritability <- read_tsv(paste0(der_dir, 'heritability.tsv'))
fluxes <- read_tsv(paste0(der_dir, 'fluxes.tsv'))

load(file = paste0(der_dir, 'decontam.RData'))

# CH4 models
load(file = paste0(der_dir, 'response_model.RData'))
load(file = paste0(der_dir, 'response_plot.RData'))
load(file = paste0(der_dir, 'herit_model.RData'))
load(file = paste0(der_dir, 'herit_plot.RData'))

# 16S models
load(file = paste0(der_dir, 'richness_models.RData'))
load(file = paste0(der_dir, 'phyloseq.RData'))
load(file = paste0(der_dir, 'beta_model.RData'))
load(file = paste0(der_dir, 'beta_fig.RData'))
load(file = paste0(der_dir, 'da_corncob.RData'))
load(file = paste0(der_dir, 'da_fig.RData'))
load(file = paste0(der_dir, 'dbrda.RData'))

@@

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know a priori which
members of a microbiome influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per passage. In addition, we
estimated that `r myround((coefficients(herit_model)[2] +
coefficients(herit_model)[4])*100, 1)`\% of the variation in methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; rather microbes that
responded to selection included members of `r length(da_fams)` families not
known to oxidize methane, such as
*Fimbriimonadaceae*, *Cytophagaceae*, and *Diplorickettsiaceae*. This result is
in contrast to the typical assumption that the rate of an ecosystem function is
limited by the final step in the associated microbial pathway. Our study
demonstrates that variation in microbiome composition can contribute to
variation in the rate of ecosystem function and that manipulating microbiome composition
could be a viable strategy for managing ecosystem functions.

@@

Microbiomes mediate a variety of important functions in ecosystems, and there
is great interest in understanding how attributes of microbiomes may influence
variation in ecosystem functions [@conrad1996; @schimel1998; @crowther2019].
Biodiversity-ecosystem function relationships have been documented for a
variety of macroorganismal communities. For example, plant species richness
correlates with productivity and marine community diversity correlates with
production, consumption, and nutrient release 
[@cardinale2006; @hooper2012; @gamfeldt2015]. However, we have been unsuccessful
in documenting similar relationships for microbial communities. While there is
some evidence that microbiomes from different ecosystems display different
functional rates [@balser2005; @strickland2009], adding microbiome attributes to
ecosystem models adds very little to total variance explained [@graham2016].
Therefore, we still lack a detailed understanding of what
functions are altered by microbiome composition and what attributes of the
microbiome explain this relationship.

There are several reasons for this difficulty. One reason is that it is
challenging to manipulate microbiome composition and diversity directly.
Experimental manipulations of community composition and diversity have been
central to documenting the relationships between ecosystem functions and the
biodiversity of plants and animals [@hooper2012], but this is seldom possible
with microbiomes. Most microorganisms cannot yet be cultured and most available
methods for manipulation of microbiomes (e.g., antibiotics, dilution, etc.) are
not sufficiently specific to manipulate individual microbial lineages. In
addition, many studies have looked for relationships between an ecosystem
function and the abundance or diversity of a microbial gene that codes for an
enzyme or enzyme-subunit known to be involved in that function. However, it is
difficult to identify the “right” marker gene a priori and for most functions
in most ecosystems there is no correlation between the abundance of a marker
gene and the rate of the corresponding process [@rocca2015]. Lastly,
microbiome-ecosystem function relationships have been difficult to document
because microbiomes often covary with other drivers of variation in ecosystem
functions, such as environmental conditions or resources. Such covariation can
make it impossible to determine whether microbiomes directly contribute to
functional variation. Understanding the degree to which microbiomes directly
influence functional variation is crucial if scientists are to develop
microbiome-focused methods for altering or maintaining ecosystem functions.

One way to overcome these challenges is to manipulate the relationship between
microbiome biodiversity and ecosystem function using artificial ecosystem selection 
[@swenson2000; @blouin2015; @panke-buisse2015; @sanchez2021]. 
Microbiomes have been shown to respond to selection at the ecosystem-trait
level [@swenson2000], and some ecosystem-scale microbiome traits have been
shown to be transferrable through microbiome inoculations [@panke-buisse2015].
Therefore, applying artificial selection to whole microbiomes could be a novel
approach for describing a more detailed microbiome-ecosystem function
relationship. This approach works by subsampling a microbiome to inoculate
multiple replicate ecosystems that vary in microbiome composition and then
measuring the rate of ecosystem function for each replicate. The investigator
then selects the replicates with the highest (or lowest) rates and subsamples
those ecosystems to passage them to a new set of ecosystems. By holding
environmental conditions relatively constant among the replicates and from
passage to passage (or by using isogenic hosts, in the case of host-associated
microbiomes), investigators can determine whether variation in the microbiome
contributes to variation in the rate of ecosystem function independent of the
environment.

A valuable feature of this approach is that it can be used to estimate how much
of the variation in ecosystem function is due to microbiome variation.  By
calculating the slope of the regression between the "donor" ecosystems selected
in one passage and the "recipient" ecosystems inoculated by those donors in the
next passage, we can quantify an estimate of "ecosystem heritability"
[@goodnight2000; @blouin2015]. The concept of heritability has been borrowed
from quantitative genetics where heritability is defined as *the ratio of
additive genetic variance to the total phenotypic variance* and is commonly
estimated as *the slope of the linear regression between offspring phenotype
and parental phenotype* [@bell1977; @falconer1996]. Here we apply this concept
to donor and recipient communities in an artificial ecosystem selection
experiment to estimate *the variance in ecosystem function due to microbiome
variance*. We do not assume that environmental variation is zero, instead we
quantify the proportion of variation due to microbiome effects. If the slope of
the regression is zero, we would conclude that environmental variation among
the replicates and technical error are the primary drivers of variation in
ecosystem function. On the other hand, a significant slope between donors and
recipients would demonstrate that microbiome variation generates variation in
the rate of an ecosystem function. We then use our artificially selected
microbiomes to investigate what aspects of microbiome variation explain this
response. By "microbiome variation" we mean aspects of microbial diversity
commonly measured by microbial ecologists, such as the relative abundance of
taxonomic groups inferred from amplicon sequence variants. With this approach,
we can not only determine whether an ecosystem function is influenced by
microbiome variation, but also estimate how much of the variation and which
taxa are associated with higher rates of ecosystem function.

We applied artificial ecosystem selection to soil microbial communities by
selecting on soil methane oxidation rate. We chose this function because
methane is a globally important greenhouse gas and methane oxidation by soil
bacteria is the primary biological sink for atmospheric methane
[@kirschke2013]. In addition, there is evidence that soil methane oxidation
rate may vary with microbiome composition. For example, soil microbiome
phylogenetic variation is a strong predictor of methane oxidation in forests
and pastures of the Brazilian Amazon as well as soils of the Congo Basin
[@meyer2020; @meyer2020a]. Finally, methanotrophy is one of the most deeply
conserved microbial physiologies and is represented in a narrow range of taxa
and so the taxonomic composition of the microbiome is more likely to be
associated with the rate of methane oxidation than other broader or more
shallowly conserved functions [@schimel1998; @martiny2013]. This suggests that
microbiome variation might drive variation in methane oxidation in soils. One
other study has estimated ecosystem heritability for an environmental
microbiome. Blouin et al. [@blouin2015] selected for water microbiomes with
low carbon dioxide respiration rates and estimated a heritability of 52%.
However, this appeared to be driven primarily by biomass, not microbiome
composition. Here we want to build on this work by selecting for higher rates
of a more narrow function that is less likely to be driven by total biomass.
That way we can begin to assess whether variation in microbiome composition is
important for understanding variation in ecosystem functions.

In this study, we evaluated whether variation in the microbiome contributes to
variation in the rate of ecosystem functions. To address this problem, we
applied artificial ecosystem selection to methane oxidation rate of the soil
microbiome and answered the following questions: Does variation in the relative
abundance of microbial taxa contribute to variation in soil methane oxidation
rate in the lab? How much of the variation in soil methane oxidation rate can
be attributed to the microbiome? What attributes of the microbiome regulate
variation in methane oxidation rate? And do these attributes match our
assumptions about what regulates methane oxidation rate in nature?

@@

We performed an artificial ecosystem selection experiment 
[*sensu* @swenson2000] by passaging replicate soil microbiomes. The trait we
selected on was CH~4~ oxidation rate. This experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were homogenized to inoculate the next set of positive jars. The
other line was a neutral selection line where an equal number of jars as the
positive line were chosen at random to inoculate the next set of neutral jars.
The number of jars chosen was based on the distribution of fluxes among the
positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater CH~4~ oxidation rates based on visual inspection of
histograms. The experiment was carried out over five passages until a
significant divergence in functional rates was observed between the two
selection treatments based on a difference of slopes.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting
mix was Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). Each jar
was then capped and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the positive or neutral selection line. Jars were
flushed and respiked with methane twice per week to maintain aerobic conditions and elevated
CH~4~ concentrations and were incubated at ambient temperature for
approximately three weeks. Methane oxidation rates were determined at the end
of the incubation period. For the positive treatment, the three jars with the
greatest CH~4~ oxidation rates were chosen to inoculate the next generation.
For the neutral treatment, three jars were randomly selected to inoculate the
next generation. For each treatment, the selected jars were homogenized and 5 g
of the homogenized soil was used to inoculate the next set of jars, which
represents a 10% subsampling. The next set of jars were created in an identical
manner to the first generation with fresh autoclaved potting mix and the same
moisture and CH~4~ content.

@@

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from
each jar immediately after spiking and then at time points 3, 6, 24, and 48
hours for a 5-point curve. Samples were immediately injected into a SRI model
8610C gas chromatograph equipped with a flame ionization detector (SRI
Instruments, Torrance, CA, USA) to determine the headspace CH~4~ concentration.
We applied a first-order exponential decay function to determine the rate
constant (k, units = d^−1^; i.e., dCH~4~/dt = k[CH~4~]) of the exponential
decrease in methane. Oxidation rates are presented as the additive inverse of
$k$ (i.e., $-k$) so that a more positive value represents a greater
oxidation rate. The jars selected in passage two for the positive treatment had
the lowest CH~4~ oxidation rate of the twelve jars due to a calculation error
in the rate constant. Despite this error, we still observed a strong
response to selection demonstrating the power of this technique. In addition,
this likely reintroduced diversity lost during the selection process,
potentially increasing the variation available to be selected upon, as
is recommended for microbiome selection experiments [@sanchez2021].

@@

Statistical analyses were performed in 'R' [@rcoreteam2018]. To determine
whether there was a significant change in CH~4~ oxidation rate as a response to
selection, we tested a difference in slopes between the positive and neutral
selection lines. Residuals did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed prior to analysis. First, we tested if there was a difference of
slopes between the positive and neutral treatments based on the interaction
between passage and treatment. To test the interaction, we fit two nested models
with and without the interaction term and compared them using an F-test with the
'anova' function. We then present the slopes for each treatment, which
represented the change in CH~4~ oxidation rate per passage as a response to
selection.

We estimated the ecosystem heritability of methane oxidation rate, which is the
proportion of variance in CH~4~ oxidation rate due to variation in the
microbiome, as the regression of mid-recipient on mid-donor [@falconer1996].
The mid-donor was the mean for all three selected jars and the mid-recipient was
the mean methane oxidation rate for the twelve jars produced by those three
donors. First, we tested if there was a difference of slopes between the
positive and neutral treatments. To test this, we fit two nested models with and
without the treatment term and compared them using an F-test with the 'anova'
function. We then present the slopes for each treatment, which represented the
ecosystem heritability or the proprotion of total variance in methane oxidation
rate due to microbiome variance.

First, we tested if there was an effect of treatment by comparing
nested models with and without treatment using an F-test with the 'anova'
function. We then fit the full model to estimate the slope of the relationship
between recipient and donor for the Positive and Neutral treatments.

@@

To identify taxa that responded to selection on CH~4~ oxidation rate, we tested
differential abundance between the two treatments in passage 5. We first
grouped ASVs at the family level. Any ASVs that lacked a family-level
taxonomic assignment were grouped at a higher taxonomic level. We then subset the
samples in Passage 5 and removed all families with a prevalence of less than
10\% in either treatment. We used three methods for testing differential
abundance: ANCOM-II, ALDEx2, and CORNCOB 
[@kaul2017; @lin2020; @fernandes2014; @martin2020]. We then identified the
consensus taxa that were significant with all three tests and plotted their
relative abundances. For ANCOM-II, we used the 'ancom' function in the
'ANCOM-BC' package with a cutoff of W = 0.7 [@kaul2017; @lin2020]. For ALDEx2,
we used the 'aldex' function in the 'ALDEx2' package with Welch's t-test and we
used an effect size of 1 as our significance threshold [@fernandes2014].
Finally, we used CORNCOB with the 'differentialTest' function in the 'corncob'
package with the Wald test and without bootstrapping [@martin2020]. Lastly, to
test differentially abundant methanotrophs, we subset all ASVs within
methanotrophic families and tested their differential abundance aggregated at
the family and genus level using 'corncob'.

@@

We observed a response to artificial selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`; difference of slopes:
F~`r compare_response_mods[2,4]`,`r compare_response_mods[2,2]`~ = `r myround(compare_response_mods[2,5], 2)`, p = `r myround(compare_response_mods[2,6], 2)`). At the start of the experiment, the Positive
treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\% lower than the
Neutral treatment `r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate of the Neutral
treatment over the five passages `r print_lm(response_model, 2)`. By contrast,
the Positive treatment had a `r myround(per_slope, 1)`\% increase in CH~4~
oxidation rate per passage `r print_lm(response_model, 4)`.

We estimated the ecosystem heritability defined as the 
proportion of variation in methane oxidation due rate to variation
in the microbiome as the regression of mid-recipient on mid-donor 
(`r figures("herit", display = "cite")`).  Recipient CH~4~ oxidation rates were
correlated with donor CH~4~ oxidation rates in both the Positive treatment
`r paste0('(', 'slope', ' = ', myround(tidy(herit_model)[2, 2] + tidy(herit_model)[4, 2], 2), ', SE = ', myround(tidy(herit_model)[4, 3], 2), ', t = ', myround(tidy(herit_model)[4, 4], 2), ', p = ', myround(tidy(herit_model)[4, 5], 2), ')')` and 
the Neutral treatment `r print_lm(herit_model, 2)`.

@@

While it is well understood that the microbiome mediates a variety of ecosystem
functions, a fundamental question in microbial ecology is whether we need to
understand variation in the composition of microbiomes to understand variation
in the rate of ecosystem functions [@conrad1996; @schimel1998]. To address this
question, we performed artificial ecosystem selection on soil microbiomes by
selecting for ecosystems with high methane oxidation rates. We found that there
was a significant increase in methane oxidation rate in the Positive selection
treatment relative to the Neutral selection treatment. This demonstrates that
passaging the microbiome through artificial ecosystem selection resulted in a
shift in the mean methane oxidation rate of the soil microbiome. 
This suggests that variation in the microbiome can generate variation in
ecosystem functions independent of the environment.
